Syntax Is All You Need: A Universal-Language Approach to Mutant Generation

### 2023-09-21 paper/main.tex
Updating comby numbers
Adding GPT-4 + Wodel

@@

% ---------------- C++ table (UM vs Others)--------------------------
\begin{table}[htbp]
\centering
{\small
\caption{C++ (Our Implementation vs. Dextool). We cannot measure equivalent mutants for Dextool.}
\label{tab:table_cpp2}
\begin{tabularx}{\columnwidth}{X|rrr|rr}
\toprule
        & \multicolumn{3}{c|}{\textbf{Comby}}  & \multicolumn{2}{c}{\textbf{Dextool}} \\ 
\textbf{Project} & \textbf{Muts.} & \makecell{\textbf{Mut.} \\ \textbf{Score}} & \textbf{\% Eqv.} & \textbf{Muts.} & \makecell{\textbf{Mut.} \\ \textbf{Score}} \\[1ex]\midrule
 ompl &  564 & 0.07 & 26.7\% & 278 & 0.08  \\
 libgraphqlparser  &  131 & 0.90 & 33.3\% & 52 & 0.96  \\
 rethinkdb\_rebirth &  492 & 0.10 & 12.7\% & 25 & 0.36 \\
 xoreos &  2280 & 0.01 & 3.3\% & 414 & 0.02  \\
 libxmljs &  790 & 0.64 & 6.7\% & 830 & 0.46 \\[0.5ex]
 tiny-differentiable-simulator &  770 & 0.04 & 0\% & 61 & 0.21  \\\midrule
\textbf{Mean}   & 837.83 & 0.29 & 16.67\% & 276.67 & 0.33\\
\textbf{Std.Dev.} & 745.78 & 0.37 & 13.66\%   & 311.56 & 0.38 \\\bottomrule
\end{tabularx}}
\end{table}

@@

@@

Hariri et al. compared C mutation approaches at the source and
compiler IR levels~\cite{CompareSrcBinary} and found that overall
source level mutation was better, producing fewer mutants overall, but
matching the IR approach in the important measures of surface and
minimal mutants and overall mutation score.  Numerous studies compare
Java mutation tools~\cite{MajorPIT,gopinath2017does}, including a
recent article for a more general audience in Communications of the
ACM~\cite{CommACMJavaTool} (perhaps showing the growing interest in
practical mutation testing).  This paper showed that users
considered active maintenance, support for a variety of testing
frameworks, and support for recent Java versions as the most important
features in a Java mutation tool.  The approach proposed in this paper
by its nature tends to promise all three of these key factors without
imposing burdens on maintainers.

>>>>

\lstdefinelanguage{haskell}{
  morekeywords={import,where,let,in,do,case,of,if,then,else,module,mutateAndTest,ParseFailed,ParseOk,mutateModule,mutateExp,main},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[n]{\{-}{-\}},
  morestring=[b]",
  basicstyle=\small\ttfamily,
  keywordstyle=\color{blue},
  commentstyle=\color{olive},
  stringstyle=\color{red},
  breaklines=true,
  breakatwhitespace=true,
  literate={->}{$\rightarrow$}{1},
}

@@

% ---------------- C++ table (UM vs Others)--------------------------
\begin{table}[htbp]
\centering
{\small
\caption{C++ (Our Implementation vs. Dextool). We cannot measure equivalent mutants for Dextool.}
\label{tab:table_cpp2}
\begin{tabularx}{\columnwidth}{X|rrr|rr}
\toprule
        & \multicolumn{3}{c|}{\textbf{Comby}}  & \multicolumn{2}{c}{\textbf{Dextool}} \\ 
\textbf{Project} & \textbf{Muts.} & \makecell{\textbf{Mut.} \\ \textbf{Score}} & \textbf{\% Eqv.} & \textbf{Muts.} & \makecell{\textbf{Mut.} \\ \textbf{Score}} \\[1ex]\midrule
 ompl &  564 & 0.07 & 26.7\% & 278 & 0.08  \\
 libgraphqlparser  &  131 & 0.90 & 33.3\% & 52 & 0.96  \\
 rethinkdb\_rebirth &  492 & 0.10 & 12.7\% & 25 & 0.36 \\
 xoreos &  2280 & 0.01 & 3.3\% & 414 & 0.02  \\
 libxmljs &  790 & 0.64 & 6.7\% & 830 & 0.46 \\[0.5ex]
 tiny-differentiable-simulator &  770 & 0.04 & 0\% & 61 & 0.21  \\\midrule
\textbf{Mean}   & 837.83 & 0.29 & 13.67\% & 276.67 & 0.33\\
\textbf{Std.Dev.} & 745.78 & 0.37 & 13.50\%   & 311.56 & 0.38 \\\bottomrule
\end{tabularx}}
\end{table}

@@

\subsection{Comparison with GPT 4}

\begin{lstlisting}[language=haskell,caption={Haskell Mutation Tool},label=lst:mutationTool]
{-# LANGUAGE FlexibleContexts #-}

module MutationTool where

import Language.Haskell.Exts

-- Main function
mutateAndTest :: String -> IO ()
mutateAndTest filePath = do
    parsedModule <- parseFile filePath
    case parsedModule of
        ParseFailed srcLoc err -> print err
        ParseOk modul -> do
            let mutatedModule = mutateModule modul
            writeFile "Mutated.hs" $ prettyPrint mutatedModule
            -- Here, you'd typically run your tests on "Mutated.hs"
            -- For now, just printing the mutated code:
            putStrLn $ prettyPrint mutatedModule

-- Perform mutation on a module
mutateModule :: Module -> Module
mutateModule = everywhere (mkT mutateExp)

-- Perform mutation on an expression
mutateExp :: Exp -> Exp
mutateExp (If cond trueBranch falseBranch) = If cond falseBranch trueBranch
mutateExp x = x

main :: IO ()
main = mutateAndTest "YourFile.hs"
\end{lstlisting}

\begin{lstlisting}[caption={GPT-4 Comby Haskell Rules},label=lst:haskellcomby]
head :[list] ==> last :[list]
tail :[list] ==> init :[list]
(:[expr1]) && (:[expr2]) ==> (:[expr1]) || (:[expr2])
(:[expr1]) || (:[expr2]) ==> (:[expr1]) && (:[expr2])
(:[expr1]) >>= (:[expr2]) ==> (:[expr1]) >> (:[expr2])
...
\end{lstlisting}

To understand how our approach compares with state of the art large language models (LLM), we prompt GPT-4 to generate a mutation testing tool for Haskell.
Surprisingly, the generated program in \CRef{lst:mutationTool} has no syntax errors, and even runs to produce a mutant. However, the program only implements 
the inverting conditionals operation, a small subset of all the operators our tool implements. Furthermore, like existing mutation testing tools for source code,
the generated code performs AST manipulation in Haskell, making it harder to read than the comby rules of our approach. Asking the LLM to expand the set of rules,
 produces more operators, however the number of mutants generated by such an approach would be a small subset of what our approach is capable of doing.

While LLMs are not able to generate fully working language specific mutation testing tools from scratch, they are capable of generating many interesting rules
 that are specific to a given language. For example, prompting GPT-4 to generate Haskell rules generates the set of rules seen in \Cref{haskellcomby}. The generated rules are 
 both valid comby syntax and language specific. Specifically, the replacing head with tail is a rule that would only apply to functional programming languages such as 
 Haskell.

@@

Hariri et al. compared C mutation approaches at the source and
compiler IR levels~\cite{CompareSrcBinary} and found that overall
source level mutation was better, producing fewer mutants overall, but
matching the IR approach in the important measures of surface and
minimal mutants and overall mutation score.  Numerous studies compare
Java mutation tools~\cite{MajorPIT,gopinath2017does}, including a
recent article for a more general audience in Communications of the
ACM~\cite{CommACMJavaTool} (perhaps showing the growing interest in
practical mutation testing).  This paper showed that users
considered active maintenance, support for a variety of testing
frameworks, and support for recent Java versions as the most important
features in a Java mutation tool.  The approach proposed in this paper
by its nature tends to promise all three of these key factors without
imposing burdens on maintainers. More recently, there has been some effort to create a 
language agnostic tool~\cite{wodeltest}. However, their tool uses Wodel, a DSL, which 
is required for the user to define new operators. Unlike comby, these operators are much harder to read, thus 
making their tool less maintainable.
